The Bolocam Galactic Plane Survey: Survey Description and Data Reduction

### 2009-04-11 methods/ms.tex
Yee haw!

% -----------------------------------------------------------------------------
% BGPS specific definitions
\def\bgps{BGPS}
\newcommand{\bgpsarea}{150}
\newcommand{\bcamfwhm}{31.2\arcsec}
\newcommand{\ncores}{7628}
\newcommand{\bgpsdepthlow}{20}
\newcommand{\bgpsdepthhigh}{50}
% -----------------------------------------------------------------------------

@@

The advent of focal plane arrays containing hundreds of individual
bolometers sensitive to millimeter and sub-millimeter (sub-mm)
radiation has enabled large-scale surveys of mm-wavelength continuum
emission from Galactic interstellar dust.  The great advantage of
continuum observations at 1.1mm is the low optical depth of the dust,
permitting an estimate of the mass of the emitting region.  Further,
the mass column density is independent of the gas volume density.

Thus it is thus possible with a wide area, relatively shallow 1.1~mm
survey to provide an unmatched, uniform inventory of massive
star-forming and starless condensations.

This survey presents the results of one of the first wide-area,
untargeted surveys of the Galactic Plane in the millimeter continuum.
We have used the 144 element Bolocam focal plane array, mounted at the
Cassegrain focus of the 10 meter diameter sub-mm telescope at the
Caltech Sub-millimeter Observatory (CSO), to survey over \bgpsarea\
square degrees of the northern Galactic plane at a wavelength of 1.1
mm in the dust continuum.  Figure \ref{fig:Coverage} shows the
coverage of the \bgps.  Maps of the entire BGPS are presented in
Figure \ref{fig:BGPSMontage}.  We have detected \ncores\ clumps in the
surveyed fields, providing an unbiased sample of clumps suitable for
multi-wavelength and high-resolution studies with existing telescopes
and future facilities.  The clump catalog is described in a companion
paper \citep{rosolowsky09}.

At the distances that we are observing these objects, we are likely
observing aggregates of star forming cores found in a larger molecular
cloud environment, which we will refer to as clumps
\citep{williams00}.

The outline of the paper is as follows. Section \ref{sec:Observations}
describes the instrument and the observations.  Section
\ref{sec:FluxCalibration} describes the flux calibration and errors.
Section \ref{sec:Astrometry} describes the construction of an absolute
reference frame for the observations and verification against other
(sub)millimeter catalogs.  Section \ref{sec:Mapping}, \ref{sec:Noise},
\ref{sec:Photometry}, \ref{sec:FinalMaps}, \ref{sec:Discussion}.

@@

%\begin{sidewaysfigure}[h]
\begin{figure}
  \begin{minipage}{6.5in}
    \begin{center}
 %     \includegraphics[scale=0.9]{fig1_pg1_bw_inv_bar_label}
      \caption{$l=-10.5$ to $l=19.5$ Units are Jy/Beam.  The brightest
	sources, e.g. Sgr B2, Sgr A, and sources near $l=10$ and
	$l=13$, appear to be saturated, but this is only a display
	artifact.  The noise is more pronounced from $l=-7$ to $l=-2$
	because this region was observed less.}
    \end{center}
    \label{fig:BGPSMontage}
  \end{minipage}
\end{figure}

@@

\begin{figure}
  \begin{minipage}{6.5in} 
    \begin{center}
 %     \includegraphics[scale=0.9]{fig1_pg2_bw_inv_bar_label} 
      \caption{$l=19.5$ to $l=49.5$.  Units are Jy/Beam .  G34.3+0.15, W
	51, W 43, W 49, and M 17 appear to be saturated, but this is
	only a display artifact}
    \end{center}
  \end{minipage}
\end{figure}

@@

\begin{figure}
  \begin{minipage}{6.5in} 
    \begin{center}
%      \includegraphics[scale=0.9]{fig1_pg3_bw_inv_bar_label}
      \caption{$l=49.5$ to $l=74.5$.  \TBD\ Add annotations}
    \end{center}
  \end{minipage}
\end{figure}

@@

\begin{figure}
  \hspace{-1in}
%  \includegraphics[scale=0.8]{fig1_pg5_bw_inv_bar_label} 
  \caption{The Cygnus Arm. \TBD\ Add annotations}
\end{figure}
%  \vspace{-4in}

@@

\begin{figure}
\hspace{-0.5in}
%  \includegraphics[scale=0.8]{fig1_pg4_bw_inv_bar_label} 
  \caption{Cloud complexes centered at $l=111$ in the Perseus Arm. \TBD\ Add annotations}
\end{figure}

@@

\begin{figure}
  \hspace{-1in}
%  \includegraphics[scale=0.8]{fig1_pg6_bw_inv_bar_label} 
  \caption{W3/4/5 \TBD\ Add annotations}
\end{figure}

@@

%\include{bgps_methods_observations}
\section{Observations}
\label{sec:Observations}

%\subsection{Instrument}

@@

%{\it material from the Letter here}

@@

The data were reduced with a custom pipeline written specifically for
the galactic plane survey based on \citet{enoch06} and
\citet{Laurent2006}.  A key element of the reduction is the iterative
estimation of the atmospheric fluctuations and the astrophysical
signal.  The atmospheric model is developed from a set of principal
components of the bolometer signal time series under the assumption
that the bulk of the correlated signal is atmospheric.  The iterative
estimation process then helps to restore correlated flux due to
astrophysical emission \citep[details in][]{Aguirre2009}.

The Bolocam observations are sensitive to structure between 31\arcsec\
and about 400\arcsec\ (the array field of view).  An indicative
analysis of the spatial filter function of the iterative mapping
scheme is given in \citet{enoch06}.  The data were sampled onto a
uniform 7\farcs2 grid.  Flux density calibrations were performed using
observations of Mars, Uranus, and Neptune, and they are accurate to
$\sim$20\%.

@@

The absolute pointing accuracy of the survey has been measured to be
$\sim$6\arcsec\ RMS.  A single $3^\circ \times 1^\circ$ observation of
each region was registered to the pointing calibrators as a `master
field,' and additional observations of that field were aligned to the
master.

Bright and large sources leave an impact on their surroundings when
cleaned with a principal component subtraction.  \object{Sgr B2},
\object{G 34.3+0.15}, and other sources with fluxes $\geq 10$ Jy are
surrounded by substantial negative bowls.  Some flux is restored in
these bowls using the iterative process described in
\citet{Aguirre2009}, but around the brightest sources the impact of
the source on the background estimation is too great to be fully
recovered.

Instrument glitches and cosmic ray hits have been removed by a
combination of outlier rejection in each spatial pixel and hand
flagging of the timestream data.  Streaking in the images results from
incorrect determinations of the zero level across scans when a bright
source is encountered.  Atmospheric noise in observations taken in
poor weather may not be completely removed by the reduction
procedures.
%Particular bolometers may be noisy for a single observation and they may not
%be adequately down-weighted.

@@

Further details of the pointing calculation are given in Appendix
\ref{sec:PointingCalculation}.

@@

We proceed iteratively, estimating $a$ and $e$ in the order of their
relative strengths, beginning with the atmosphere.  The first
iteration step is then
\begin{equation}
\label{eq:FirstIterStep}
\tilde{a}^{(n)}(t) = 
\sum_i{d_i(t)-\tilde{s}^{(n)}_i(t) - \tilde{e}^{(n)}_i(t)}
\end{equation}
where $\tilde{s}^{(0)}_i(t)=0$.  This mean atmosphere model is then
fit to each bolometer to obtain the relative gains (flat field) of the
detectors
\begin{equation}
\tilde{a}^{(n)}_i(t) = \frac{d_i(t) \cdot a(t)}{d_i(t)^2} a(t) = r_i a(t)
\end{equation}
This is illustrated in Figure \ref{fig:Flatfield}.  The instrument
errors, which are typically smaller than the atmosphere, are then
estimated
\begin{equation}
\tilde{e}^{(n)}_i(t) = some models
\end{equation}
and both the atmosphere and instrument models are subtracted from the
raw time series
\begin{equation}
d_i(t) - \tilde{a}^{(n)}_i(t) - \tilde{e}^{(n)}_i(t)
\approx s_i(t) + \epsi_i(t)
\end{equation}
This is the best estimate of signal plus irreducible noise at
iteration step $n$, and is made into a map
\begin{equation}
M[d_i(t) - \tilde{a}^{(n)}_i(t) - \tilde{e}^{(n)}_i(t)] = \tilde{m}^{(n)}
\end{equation}
The current best map $\tilde{m}^{(n)}$ is then deconvolved to provide
a relatively low-noise, smooth map from which to generate a timestream
\begin{equation}
T[{\cal D}[\tilde{m}^{(n)}]] = \tilde{s}^{(n)}_i(t)
\end{equation}
where ${\cal D}$ represents the deconvolution operation.  At this
stage, the iterative process could begin again with Equation
\ref{eq:FirstIterStep}, though it is highly useful to produce an
``error map''
\begin{equation}
E = M[d_i(t) - \tilde{s}^{(n)}_i(t) - \tilde{a}^{(n)}_i(t) - 
\tilde{e}^{(n)}_i(t)]
\end{equation}
This error map serves as a straightforward visualization of the
progress of the iterative method, since in a perfect process it would
be map of $\epsi$.

@@

%WRITTEN POORLY: 
The timestream data is made into a spatial map using the pointing data
corresponding to each time point for each bolometer.  The data is
weighted by inverse variance across a single scan and then drizzled
into a map with $7.2$\arcsec\ pixels using a nearest-neighbor
algorithm.  The nearest-neighbor matching allows the map to be
returned to a timestream in the same manner, but with the S/N improved
by averaging over all hits on a given pixel.  It has the disadvantage
that it accentuates the unevenness of the cross-scan sampling, since
ech timestream point's value is assigned to an area much smaller than
the beam.

The resulting map is then subjected to a Hogbom-type CLEAN algorithm
\citep{hogbom} using a specified kernel to produce a deconvolved
image.  This has several advantages over the

@@

% this is not true.  I can't do jackknifing....
% To estimate the size of this
% residual error, we also construct jackknife maps for each field, in
% which the astrophysical signal is made to exactly cancel.  This
% provides an estimate of the noise-only RMS in the maps, to which
% deviations from the ``ghosts'' may be compared.

The noise is correlated between adjacent pixels.

The final per-pixel error estimate is given as ...

@@

@@

The final maps are produced by coadding all observations of a single
square degree centered on a particular Galactic coordinate
$(l_0,b_0)$.  The maps are made in Galactic coordinates using a plate
carr\'{e} (FITS header CAR) projection.  As these maps are near the
coordinate system equator, the difference between truly equiareal
pixels and the pixels used is at most X\%.  The pixel size is
7.2\arcsec, chosen to be much smaller than the Bolocam beam, and
producing data squares 500 pixels on a side.

@@

All processed maps are made available through
IPAC\footnote{http://irsa.ipac.caltech.edu/Missions/bolocam.html}.
IPAC provides a cutout service for the

@@

% first millimeter....
The BGPS is the first (sub)millimeter survey of a substantial fraction
of the Galactic Plane, providing an unbiased look at the high-density
gas most intimately associated with the earliest phases of star
formation.  Inspection of Figure \ref{fig:BGPSMontage}

Figures 1 and 2 show the results of the BGPS.  Our images show
excellent correspondence with SCUBA maps of small sub-fields
throughout the plane \citep{SCUBAlegacy}.  BGPS contours overlaid on
the GLIMPSE survey \citep{benjamin03} show excellent correlation with
8\mum\ infrared dark clouds (IRDCs).  However, many features in the
BGPS maps do not appear in the Spitzer observations, which suggests
that the 1.1 mm emission traces dense clumps without a background to
absorb.  A study of IRDC correlation with BGPS emission is underway
\citep{Battersby2009}.

@@

@@

\section{Pointing Calculation}
\label{sec:PointingCalculation}

>>>>

% -----------------------------------------------------------------------------
% BGPS specific definitions
\def\bgps{BGPS}
\newcommand{\bgpsarea}{150}
\newcommand{\bcamfwhm}{31.2\arcsec}
\newcommand{\ncores}{7628}
\newcommand{\bgpsdepthlow}{20}
\newcommand{\bgpsdepthhigh}{50}
% -----------------------------------------------------------------------------
% Missions
\def\spitzer{{\em Spitzer}}
% -----------------------------------------------------------------------------

@@

The advent of focal plane arrays containing hundreds of individual
bolometers sensitive to millimeter and sub-millimeter (sub-mm)
radiation has enabled large-scale surveys of mm-wavelength continuum
emission from Galactic interstellar dust.  The great advantage of
continuum observations at 1.1mm is the low optical depth of the dust,
permitting an estimate of the mass of the emitting region.  Further,
the mass column density is independent of the gas volume density.
Thus it is thus possible with a wide area, relatively shallow 1.1~mm
survey to provide an unmatched, uniform inventory of massive
star-forming and starless condensations.

This survey presents the results of one of the first wide-area,
untargeted surveys of the Galactic Plane in the millimeter continuum.
We have used the 144 element Bolocam focal plane array, mounted at the
Cassegrain focus of the 10 meter diameter sub-mm telescope at the
Caltech Sub-millimeter Observatory (CSO), to survey over \bgpsarea\
square degrees of the northern Galactic plane at a wavelength of 1.1
mm in the dust continuum.  Figure \ref{fig:Coverage} shows the
coverage of the \bgps.  Maps of the entire BGPS are presented in
Figure \ref{fig:BGPSMontage}.  We have detected \ncores\ clumps in the
surveyed fields, providing an unbiased sample of clumps suitable for
multi-wavelength and high-resolution studies with existing telescopes
and future facilities.  At the distances that we are observing these
objects, we are likely observing aggregates of star forming cores
found in a larger molecular cloud environment, which we will refer to
as clumps \citep{williams00}.  The clump catalog is described in a
companion paper \citep{rosolowsky09}.

The outline of the paper is as follows. Section \ref{sec:Observations}
describes the instrument and the observations.  Section
\ref{sec:Overview} briefly describes the data reduction process, while
Sections \ref{sec:FluxCalibration} to \ref{sec:FinalMaps} provide
greater detail on flux calibration, astrometry, the mapping algorithm,
noise, photometry, and the public data release.  We conclude in
Section \ref{sec:Discussion} with brief discussion of the immediate
implications of the emission revealed by the BGPS.

@@

%\begin{sidewaysfigure}[h]
\begin{figure}
  \begin{minipage}{6.5in}
    \begin{center}
      \includegraphics[scale=0.9]{fig1_pg1_bw_inv_bar_label}
      \caption{$l=-10.5$ to $l=19.5$ Units are Jy/Beam.  The brightest
	sources, e.g. Sgr B2, Sgr A, and sources near $l=10$ and
	$l=13$, appear to be saturated, but this is only a display
	artifact.  The noise is more pronounced from $l=-7$ to $l=-2$
	because this region was observed less.}
    \end{center}
    \label{fig:BGPSMontage}
  \end{minipage}
\end{figure}

@@

\begin{figure}
  \begin{minipage}{6.5in} 
    \begin{center}
      \includegraphics[scale=0.9]{fig1_pg2_bw_inv_bar_label} 
      \caption{$l=19.5$ to $l=49.5$.  Units are Jy/Beam .  G34.3+0.15, W
	51, W 43, W 49, and M 17 appear to be saturated, but this is
	only a display artifact}
    \end{center}
  \end{minipage}
\end{figure}

@@

\begin{figure}
  \begin{minipage}{6.5in} 
    \begin{center}
      \includegraphics[scale=0.9]{fig1_pg3_bw_inv_bar_label}
      \caption{$l=49.5$ to $l=74.5$.  \TBD\ Add annotations}
    \end{center}
  \end{minipage}
\end{figure}

@@

\begin{figure}
  \hspace{-1in}
  \includegraphics[scale=0.8]{fig1_pg5_bw_inv_bar_label} 
  \caption{The Cygnus Arm. \TBD\ Add annotations}
\end{figure}
%  \vspace{-4in}

@@

\begin{figure}
\hspace{-0.5in}
  \includegraphics[scale=0.8]{fig1_pg4_bw_inv_bar_label} 
  \caption{Cloud complexes centered at $l=111$ in the Perseus Arm. \TBD\ Add annotations}
\end{figure}

@@

\begin{figure}
  \hspace{-1in}
  \includegraphics[scale=0.8]{fig1_pg6_bw_inv_bar_label} 
  \caption{W3/4/5 \TBD\ Add annotations}
\end{figure}

@@

\section{Observations}
\label{sec:Observations}

@@

{\it just stole material from the Letter here - needs a bit of cleaning up}

@@

The data were reduced with a custom pipeline written specifically for
A key element of the reduction is the iterative estimation of the
atmospheric fluctuations and the astrophysical signal.  The
atmospheric model is developed from a set of principal components of
the bolometer signal time series under the assumption that the bulk of
the correlated signal is atmospheric.  The iterative estimation
process then helps to restore correlated flux due to astrophysical
emission (Section \ref{sec:Mapping}).

The Bolocam observations are sensitive to structure between 31\arcsec\
and about 400\arcsec\ (the array field of view).  An indicative
analysis of the spatial filter function of the iterative mapping
scheme is given in \citet{enoch06}.  The data were sampled onto a
uniform 7\farcs2 grid.  Flux density calibrations were performed using
observations of Mars, Uranus, and Neptune, and they are accurate to
$\sim$20\%.  This is described in \ref{sec:FluxCalibration}.

@@

The absolute pointing accuracy of the survey has been measured to be
$\sim$6\arcsec\ RMS.  A single $3^\circ \times 1^\circ$ observation of
each region was registered to the pointing calibrators as a ``master
field'', and additional observations of that field were aligned to the
master (Section \ref{sec:Astrometry}).

Bright and large sources leave an impact on their surroundings when
cleaned with a principal component subtraction.  \object{Sgr B2},
\object{G 34.3+0.15}, and other sources with fluxes $\geq 10$ Jy are
surrounded by substantial negative bowls.  Some flux is restored in
these bowls using the iterative process 
%described in \citet{Aguirre2009}
, but around the brightest sources the impact of the source on the
background estimation is too great to be fully recovered (Section
\ref{sec:Photometry}).

Instrument glitches and cosmic ray hits have been removed by a
combination of outlier rejection in each spatial pixel and hand
flagging of the timestream data.  Streaking in the images results from
incorrect determinations of the zero level across scans when a bright
source is encountered.  Atmospheric noise in observations taken in
poor weather may not be completely removed by the reduction
procedures (Section \ref{sec:Noise}).
%Particular bolometers may be noisy for a single observation and they may not
%be adequately down-weighted.

@@

Further details of the pointing calculation are given in Appendix
\ref{app:PointingCalculation}.

@@

We proceed iteratively, estimating $a$ and $e$ in the order of their
relative strengths, beginning with the atmosphere.  The first
iteration step is then
\begin{equation}
\label{eq:FirstIterStep}
\tilde{a}^{(n)}(t) = 
\sum_i{d_i(t)-\tilde{s}^{(n)}_i(t) - \tilde{e}^{(n)}_i(t)}
\end{equation}
where $\tilde{s}^{(0)}_i(t)=0$.  This mean atmosphere model is then
fit to each bolometer to obtain the relative gains (flat field) of the
detectors
\begin{equation}
\tilde{a}^{(n)}_i(t) = \frac{d_i(t) \cdot a(t)}{d_i(t)^2} a(t) = r_i a(t)
\end{equation}
This is illustrated in Figure \ref{fig:Flatfield}.  The instrument
errors, which are typically smaller than the atmosphere, are then
estimated
\begin{equation}
\tilde{e}^{(n)}_i(t) = some models
\end{equation}
and both the atmosphere and instrument models are subtracted from the
raw time series
\begin{equation}
d_i(t) - \tilde{a}^{(n)}_i(t) - \tilde{e}^{(n)}_i(t)
\approx s_i(t) + \epsi_i(t)
\end{equation}
This is the best estimate of signal plus irreducible noise at
iteration step $n$, and is made into a map
\begin{equation}
M[d_i(t) - \tilde{a}^{(n)}_i(t) - \tilde{e}^{(n)}_i(t)] = \tilde{m}^{(n)}
\end{equation}
The current best map $\tilde{m}^{(n)}$ is then deconvolved to provide
a relatively low-noise, smooth map from which to generate a timestream
\begin{equation}
\label{eq:MapToTime}
T[{\cal D}[\tilde{m}^{(n)}]] = \tilde{s}^{(n)}_i(t)
\end{equation}
where ${\cal D}$ represents the deconvolution operation.  At this
stage, the iterative process could begin again with Equation
\ref{eq:FirstIterStep}, though it is highly useful to produce an
``error map''
\begin{equation}
E = M[d_i(t) - \tilde{s}^{(n)}_i(t) - \tilde{a}^{(n)}_i(t) - 
\tilde{e}^{(n)}_i(t)]
\end{equation}
This error map serves as a straightforward visualization of the
progress of the iterative method, since in a perfect process it would
be map of $\epsi$.

@@

%WRITTEN POORLY: 
The timestream data is made into a spatial map using the pointing data
corresponding to each time point for each bolometer.  The data is
weighted by inverse variance across a single scan and then drizzled
into a map with $7.2$\arcsec\ pixels using a nearest-neighbor
algorithm.  The nearest-neighbor matching allows the map to be
returned to a timestream in the same manner, but with the S/N improved
by averaging over all hits on a given pixel.  It has the disadvantage
that it accentuates the unevenness of the cross-scan sampling, since
ech timestream point's value is assigned to an area much smaller than
the beam.  In principle this can be addressed by various gridding
algorithms, and may be a part of future reductions of the BGPS data.

The resulting map is then subjected to a Hogbom-type CLEAN algorithm
\citep{hogbom} using a specified kernel to produce a deconvolved image
which is the one actually mapped back into the timestream to be
subtracted (Equation \ref{eq:MapToTime}.  This has several advantages
over using the original map directly, including rejecting artefacts
smaller than the beam and decreasing the noise of the resulting
timestream.  It is also better than a ``threshold'' method, which
tends to produce discontinuities.

@@

The noise is correlated between adjacent spatial pixels due to
residual atmospheric noise correlated in time..  We have completed
ignored this effect in our analysis.

The noise may be estimated either from the error maps, from the
timestream PSDs of the detectors, or from the weight maps
produced. The final per-pixel error estimate is given as \TBD.

@@

To assess the effect of the mapping algorithm, we performed a series
of simulations.

@@

The final maps are produced by coadding all observations of a single
square degree centered on a particular Galactic coordinate
$(l_0,b_0)$.  The maps are made in Galactic coordinates using a plate
carr\'{e} (FITS header CAR) projection.  (This is the same projection
used by the \spitzer-GLIMPSE and BU-FCRAO GRS surveys.)  As these maps
are near the coordinate system equator, the difference between truly
equiareal pixels and the pixels used is at most X\%.  The pixel size
is 7.2\arcsec, chosen to be much smaller than the Bolocam beam.
%, and
%producing data squares 500 pixels on a side.
The maps are written to standard FITS files.  An example of the FITS
header for the BGPS images is given in Appendix \ref{app:FITS_Header}.

@@

All processed maps are made available through
IPAC\footnote{http://irsa.ipac.caltech.edu/Missions/bolocam.html}.
IPAC provides a cutout service for the images as well as a searchable
version of the catalog provided in \citep{rosolowsky09}.  The full
release inclues
\begin{enumerate}
\item Calibrated maps for all regions observed (MAP)
\item The MAP smoothed by the Bolocam PSF, which is helpful in bringing out extended structure (SMOOTHMAP)
\item The deconvolved MAP (DECONVMAP)
\item Maps of integration time per pixel (NHITSMAP)
\item Maps of noise per pixel (NOISEMAP)
\item Maps of the mask obtained for various regions from the cataloging algorithm (MASKMAP)
\end{enumerate}

@@

% first millimeter....
The BGPS is the first (sub)millimeter survey of a substantial fraction
of the Galactic Plane, providing an unbiased look at the high-density
gas most intimately associated with the earliest phases of star
formation.  Inspection of Figure \ref{fig:BGPSMontage} shows excellent
correspondence with SCUBA maps of small sub-fields throughout the
plane \citep{difrancesco08}.  BGPS contours overlaid on the GLIMPSE
survey \citep{benjamin03} show excellent correlation with 8\mum\
infrared dark clouds (IRDCs).  However, many features in the BGPS maps
do not appear in the Spitzer observations, which suggests that the 1.1
mm emission traces dense clumps without a background to absorb.  A
study of IRDC correlation with BGPS emission is underway
\citep{battersby09}.

@@

In conclusion, the BGPS is most awesome survey ever.

\noindent---------------------------------

@@

\section{Pointing Calculation}
\label{app:PointingCalculation}
